Treatment planning for HDR Brachytherapy

### 2015-03-31 Paper Seminar/classification.tex
Fixed ref errors, did some beauty work (i.e. newpage after toc)
Rechtschreibung geprüft...

In treatment planning, there are several steps to be executed when trying to classify. First, the body to be treated ("treated body") needs to be analyzed (i.e. measured in various variables). Second, these measurements build a formula providing a norm (like the euclidian norm), which then can be used to compare the bodies. Afterwards, the body with the smallest distance to the treated body is taken. If it has an optimization result stored ("stored body), this will be used during the next iteration. This aims at lower runtimes for at least an equal coverage.

@@

The four parameters descripted above are calculated in a measuring algorithm, detailed further in section \ref{classification:algorithm}. They contain two types of influence onto the norm: \textit{relative} and \textit{absolute}. \textit{relative} depicts a comparison with respect to the treated body's values of the same parameter. For example, the tumor volume size of the stored body will be divided by the tumor volume size of the treated body. In constrast, the \textit{absolute} comparison omits the division completely. This is due to division by 0, when the tumor overlaps with other body type volumes. \\

They each share a weight parameter, which sums up to 1 in total. The values shown in table \ref{tbl:classification_parameters} represent a multiplication factor within the norm. The norm itself is a sum of the different parameters multiplied with their weight, corresponding to the smallest error between the stored and treated bodies. The weights correspond to a simple consideration about how much each parameter should influence the the classification. To allow for small to medium deviations within the three first parameters (volume size, volume center and distance tumor-center), they have been chosen small. This enables a few transformations (i.e. translation of single/all volumes) of the body volumes resulting in a similar classification. Therefore, the weight for closest distance has been chosen larger, to prevent extreme transformations to distort the result. \\

Additionally to the weights of the various weights of the measurement parameters, it's possible to change the influence of the different body types (compare table \ref{tbl:classification_type_weights}). This leads to a more detailed and precise representation of the aims during the treatment planning. In this case, a high focus has been set upon not damaging the spine and the pancreas whilst treating the tumor properly.

@@

\subsection{Algorithm}\label{classification:algorithm}
The algorithm is shown in algorithm \ref{alg:classification}. It consists of a main loop, iterating through the stored body and performing the following measurements and additional tasks. \\
First, the body type counter for the current voxel is incremented (line 2). Second, the x-, y- and z-coordinate of the voxel is added to total coordinate of the body type (line 3). This aims at the calculation of the volume center. If the body type is of \textit{tumor}, then the voxel coordinate is stored for later usage, i.e. the calculation of the closest distance parameter (line 4 \& 5). Afterwards, the parameter volume center and distance tumor-center are calculated by division and euclidian norm (line 6 \& 7).\\
To calculate the closest distance between the volumes, it's necessary cross-compare the voxels. Therefore, another loop iterates through the complete body, and, within, iterates over all previously stored tumor coordinates. For each pair of voxel and tumor coordinate, the distance is calculated and, if less then before, the closest distance is updated (line 8-12).

\ForAll{$voxels\ in\ body$}{
	$count\ body\_type;$ \\
	$sum\ up\ coordinates;$\\
	\If{$body\_type == tumor$}{
		$store\ coordinate\ for\ later\ usage;$
	}
}
$volumeCenter = coordinates / body\_type;$ \\
$centerDistance = euclid\_norm (tumorCenter, volumeCenter);$ \\
\ForAll{$voxels\ in\ body$}{
	\ForAll{$voxels\ stored\ in\ tumor$}{
		$calculated\ distance (voxel, tumor);$\\
		\If{distance $\textless$ closestDistance}{
			$update\ closestDistance;$
		}
	}
}
\caption{Measurement algorithm in pseudo code}
\end{algorithm}

\subsection{Performance evalution and prospects}
Without the calculation of the closest distance in between the tumor volume and all other volumes, the performance is linear. The computing takes only a few seconds (2-4s), using a body similar to dimensions of 110x110x200. Alas, the computing time increases exponentially when performing the cross-comparison of each voxel in the body with all voxels of type \textit{tumor}. A first step of optimization has been taken when introducing a cache for the tumor voxel (already included in the algorithm above). Another step towards lower runtimes contains the implementation of a multi-threaded cross-comparison or completely multi-threaded algorithm. The multi-threaded cross-comparison has been implemented and it indicates a speed up by approximately the number of processor cores used, but has not beend thoroughly tested yet. \\
Like the multi-threading, at the time of this paper, the classification and the results of it are not properly tested. Therefore, no scientific statement can be made. However, for linear programming no influence upon the runtime has been observed. When using the genetic algorithm or simulating annealing, only a small deviation of runtime has been observed. Unluckily it is not clear, whether the deviation is caused by the standard variation of the algorithms runtime, or by the influence of classification.\\
Therefore, a major part of future work should contain the testing of the classification. Afterwards, a founded evaluation and adaptions according to the evaluation can be made.

>>>>

In treatment planning, there are several steps to be executed when trying to classify. First, the body to be treated ("treated body") needs to be analyzed (i.e. measured in various variables). Second, these measurements build a formula providing a norm (like the euclidian norm), which then can be used to compare the bodies. Afterwards, the body with the smallest distance to the treated body is taken. If it has an optimization result stored ("stored body"), this will be used during the next iteration. This aims at lower runtimes for at least an equal coverage.

@@

The four parameters described above are calculated in a measuring algorithm, detailed further in section \ref{classification:algorithm}. They contain two types of influence onto the norm: \textit{relative} and \textit{absolute}. \textit{relative} depicts a comparison with respect to the treated body's values of the same parameter. For example, the tumor volume size of the stored body will be divided by the tumor volume size of the treated body. In constrast, the \textit{absolute} comparison omits the division completely. This is due to division by 0, when the tumor overlaps with other body type volumes. \\

They each share a weight parameter, which sums up to 1 in total. The values shown in table \ref{tbl:classification_parameters} represent a multiplication factor within the norm. The norm itself is a sum of the different parameters multiplied with their weight, corresponding to the smallest error between the stored and treated bodies. The weights correspond to a simple consideration about how much each parameter should influence the classification. To allow for small to medium deviations within the three first parameters (volume size, volume center and distance tumor-center), they have been chosen small. This enables a few transformations (i.e. translation of single/all volumes) of the body volumes resulting in a similar classification. Therefore, the weight for closest distance has been chosen larger, to prevent extreme transformations to distort the result. \\

Additionally to the weights of the various weights of the measurement parameters, it's possible to change the influence of the different body types (compare table \ref{tbl:classification_type_weights}). This leads to a more detailed and precise representation of the goals during the treatment planning. In this case, a high focus has been set upon not damaging the spine and the pancreas, whilst treating the tumor properly.

@@

\subsection{Algorithm}\label{classification:algorithm}
The algorithm is shown in algorithm \ref{alg:classification}. It consists of a main loop, iterating through the stored body and performing the following measurements and additional tasks. \\
First, the body type counter for the current voxel is incremented (line 2). Second, the x-, y- and z-coordinate of the voxel is added to total coordinate of the body type (line 3). This aims at the calculation of the volume center. If the body type is of \textit{tumor}, then the voxel coordinate is stored for later usage, i.e. the calculation of the closest distance parameter (line 4 \& 5). Afterwards, the parameter volume center and distance tumor-center are calculated by division and euclidian norm (line 6 \& 7).\\
To calculate the closest distance between the volumes, it's necessary cross-compare the voxels. Therefore, another loop iterates through the complete body and within iterates over all previously stored tumor coordinates. For each pair of voxel and tumor coordinate, the distance is calculated and, if less then before, the closest distance is updated (line 8-12).

\begin{algorithm}
\ForAll{$voxels\ in\ body$}{
	$count\ body\_type;$ \\
	$sum\ up\ coordinates;$\\
	\If{$body\_type == tumor$}{
		$store\ coordinate\ for\ later\ usage;$
	}
}
$volumeCenter = coordinates / body\_type;$ \\
$centerDistance = euclid\_norm (tumorCenter, volumeCenter);$ \\
\ForAll{$voxels\ in\ body$}{
	\ForAll{$voxels\ stored\ in\ tumor$}{
		$calculated\ distance (voxel, tumor);$\\
		\If{distance $\textless$ closestDistance}{
			$update\ closestDistance;$
		}
	}
}
\caption{Measurement algorithm in pseudo code}
\label{alg:classification}
\end{algorithm}

\subsection{Performance evaluation and prospects}
Without the calculation of the closest distance in between the tumor volume and all other volumes, the performance is linear. The computation takes only a few seconds (2-4s), using a body similar to dimensions of 110x110x200. Alas, the computing time increases exponentially when performing the cross-comparison of each voxel in the body with all voxels of type \textit{tumor}. A first step of optimization has been taken when introducing a cache for the tumor voxel (already included in the algorithm above). Another step towards lower runtimes contains the implementation of a multi-threaded cross-comparison or completely multi-threaded algorithm. The multi-threaded cross-comparison has been implemented and it indicates a speed up by approximately the number of processor cores used, but has not been thoroughly tested yet. \\
Like the multi-threading, at the time of this paper, the classification and the results of it are not properly tested. Therefore, no scientific statement can be made. However, for linear programming no influence upon the runtime has been observed. When using the genetic algorithm or simulating annealing, only a small deviation of runtime has been observed. Unluckily it is not clear, whether the deviation is caused by the standard variation of the algorithms runtime, or by the influence of classification.\\
Therefore, a major part of future work should contain the testing of the classification. Afterwards, a founded evaluation and adaptions according to the evaluation can be made.

### 2015-03-31 Paper Seminar/isim_paper.tex
Rechtschreibung geprüft...

\section{Introduction}
High dose rate brachytherapy (abbreviated: hdr brachytherapy) as a conformal radiation therapy allows a much higher cancer tissue ratio, compared to other treatment approaches. During hdr brachytherapy, radioactive sources (i.e. seeds) are placed within the patient's body, in, close to and around the cancer tissue. Then, the seeds eradiate into the surrounding tissues, remaining a pre-calculated time frame (dwell time). Therefore, seed placement and dwell times are the most important factors when analyzing brachytherapy treatments \cite{brachytherapy}. \\
The focus of this article is set upon optimizing the dwell times for a fixed seed placement using computer aided algorithms, i.e. genetic algorithms and linear programming, as well as step-wise linear programming. The different algorithms will be described. Further on, they are analyzed and evaluated upon their adjustment possibilities (like parameters), subject to different treatment goals. Additionally, conclusions about the quality of the resulting treatment plans are drawn. Lastly, a possible approach to reduce runtim complexity and reuse of prior knowledge is presented.

@@

\subsection{Body model}
The body model is supplied as a three-dimensional array ("body array"), containing the type of the body cells, within a matlab file (.mat). The body array is set up as an equidistant point cloud of cells (1 unit $\widehat{=}$ 0.2cm), being transformed into 3D-array of voxels (compare algorithm \ref{model:voxel}). Besides the coordinate, a voxel contains several dosis variables (like extreme dosis, goal dosis) and the body type (normal tissue, spine, liver, pancreas and tumor tissue). \\
\begin{algorithm}[H]
maxDosis: \textit{double} [Gy] \\
minDosis: \textit{double} [Gy] \\
goalDosis: \textit{double} [Gy] \\
currentDosis: \textit{double} [Gy] \\
coordinate: \textit{Coordinate} [x: units, y: units, z: units] \\
bodyType: int
\label{model:voxel}
\caption{Properties of a voxel}
\end{algorithm}

An example body is displayed in figure \ref{model:body}. Here, the planning target volume (PTV, i.e. the tumor) and the other volumes of interest (VOI) are displayed. The VOIs might also be referred to as organs at risk (OAR). The ptv is in the middle of the figure (drawn red), surrounded by the liver (green). The yellow, cylindric volume behind the liver on the right represents the spine. At last, the pancreas is drawn as a black sphere (bottom right from the tumor).
\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{pictures/body}
\caption{Example body with tumor}
\label{model:body}
\end{figure}

Inside the body, the seeds are positioned randomly within the tumor voxels. When eradiating into the surrounding tissue, the seeds influence overlap and form a superposition. Hence, the dosis in a voxel is calculated as a sum of each seed by evaluating the dose-function up to a fixed limite, where the influence gets negligible (see chapter \ref{dose_function}).

@@

\section{Dose-Function}\label{dose_function}
The dose-distribution within the patient's body is the accumulation of all the radiation-seeds. Therefore, it is mandatory to know the radiation intensity of each seed. Equation \eqref{eq:dosefunction} shows the general approach to approximate the dose intensity function, whilst \eqref{eq:dosefunction2} is showing the used implementation of that equation.
\begin{equation}
\label{eq:dosefunction}
\dot{D}(r,\theta_{0})= S_{k}\Lambda \frac{G_{L}(r,\theta_{0})}{G_{L}(r_{0},\theta_{0})} g_{L}(r) F(r_{0},\theta_{0})
\end{equation}
\begin{equation}
\label{eq:dosefunction2}
\Rightarrow \ \dot{D}(r,\theta_{0})= S_{k}\Lambda \frac{G_{L}(r,\theta_{0})}{G_{L}(r_{0},\theta_{0})} g_{L}(r) \cdot 1
\end{equation}
\\
The following section will describe the used parameters:
\begin{description}
\item[Air-Kerma-Strength $S_{k}$]~\\
The $S_{k}$ describes the \textbf{k}inetic \textbf{e}nergy \textbf{r}eleased per unit \textbf{ma}ss called kerma. It is measured in $U$ ($1\ U =1 \ \frac{cGy \cdot m^{2}}{h} $)   
\item[Dose Rate Constant $\Lambda$]~\\
The dose rate constant is defined as the ratio of the dose rate at reference point $ P(r_{0},\theta_{0})  $ and the the air-kerma-strength $S_{k}$. Therefore $ \Lambda \ = \ \frac{\dot{D}(r_{0},\theta_{0})}{S_{k}} $.

@@

 The population is the accumulation of all individuals (see \ref{pop}. $k$ denotes the number of individuals.
\begin{equation}
 \label{eq:population}
 Population \ \ p := \begin{pmatrix}
 i_{0} \\ \vdots \\ i_{k-1} 	
\end{pmatrix}   
 \end{equation}

@@

\paragraph{Treatment Range}
\label{para:treatment range}
As it is shown in chapter \ref{dose_function}, the distribution of the radiation decreases exponentially with increasing distance. This effect could be used to reduce the number of voxel being evaluated, as the influence of voxels relatively far away is negligibly small. Algorithm \ref{alg:scalefif} shows how that information is used. The region of interest (roi) denotes the treatment range. \\

@@

\paragraph{Weights}
It seems hard to examine a specific sets of weights, that is applicable to any type of patients. The experiment has shown, that a higher weighted treatment has a much higher runtime and the convergency decreases.  Figure \ref{fig:weights} shows the difference between the goal dose and the actual dose for different weighted distributions.
\begin{figure}[hbtp]
\begin{tikzpicture}
  \begin{axis}[
  	name=one,
  	width=0.86 \columnwidth,
  	height=1.0in, 
  	scale only axis,
  	xmajorticks=false,
  	xtick={0,10,20,30,40},
  	ytick={-20,-10,...,10},
  	minor ytick={-20,-15,...,10},
  	legend style={anchor=north,at={(0.5,1.6)},font=\small},
  	legend columns=2,
    xmin=0,
    xmax=40,
  	ymax = 5,
  	ymin = -20,
    grid=both,
    y label style={at={(axis description cs:0,.5)},anchor=south},
    ylabel = {Diff. to GD [Gy]}
    ]
	\addplot [color=black,dashed,line width=1.0pt]
	table[row sep=crcr]{0 0\\
						10 -7.4061373\\
						20 -7.3660937\\
						30 -7.16107\\
						40 -7.5223036\\	};
	\addlegendentry{Bladder};

@@

\section{Linear Programming}
In Linear Programming an objective function and constraints are used to compute the optimal solution to a minimization or maximization problem, as described in [\cite{matprog}]. The problem consists of a certain number of variables and a feasible region, which is defined by the contraints. 
Equation \eqref{eq:LPMatrix} shows the standard form of a linear problem. The objective function, modeled with \textit{c},  is to be minimized, while the constraints, modeled with \textit{A} and \textit{b}, are satisfied. The used algorithm yields a vector $x$ for the optimal solution.

>>>>

\section{Introduction}
High dose rate brachytherapy (abbreviated: hdr brachytherapy) as a conformal radiation therapy allows a much higher cancer tissue ratio, compared to other treatment approaches. During hdr brachytherapy, radioactive sources (i.e. seeds) are placed within the patient's body, in, close to and around the cancer tissue. Then, the seeds eradiate into the surrounding tissues, remaining a pre-calculated time frame (dwell time). Therefore, seed placement and dwell times are the most important factors when analyzing brachytherapy treatments \cite{brachytherapy}. \\
The focus of this article is set upon optimizing the dwell times for a fixed seed placement using computer aided algorithms, i.e. genetic algorithms and linear programming, as well as step-wise linear programming. The different algorithms will be described. Further on, they are analyzed and evaluated upon their adjustment possibilities (like parameters), subject to different treatment goals. Additionally, conclusions about the quality of the resulting treatment plans are drawn. Lastly, a possible approach to reduce runtime complexity and reuse of prior knowledge is presented.

@@

\subsection{Body model}
The body model is supplied as a three-dimensional array ("body array"), containing the type of the body cells, within a matlab file (.mat). The body array is set up as an equidistant point cloud of cells (1 unit $\widehat{=}$ 0.2cm), being transformed into 3D-array of voxels (compare algorithm \ref{model:voxel}). Besides the coordinate, a voxel contains several dose variables (like extreme dose, goal dose) and the body type (normal tissue, spine, liver, pancreas and tumor tissue). \\
\begin{algorithm}[H]
maxDosis: \textit{double} [Gy] \\
minDosis: \textit{double} [Gy] \\
goalDosis: \textit{double} [Gy] \\
currentDosis: \textit{double} [Gy] \\
coordinate: \textit{Coordinate} [x: units, y: units, z: units] \\
bodyType: int
\label{model:voxel}
\caption{Properties of a voxel}
\end{algorithm}

An example body is displayed in figure \ref{model:body}. Here, the planning target volume (PTV, i.e. the tumor) and the other volumes of interest (VOI) are displayed. The VOIs might also be referred to as organs at risk (OAR). The PTV is in the middle of the figure (drawn red), surrounded by the liver (green). The yellow, cylindric volume behind the liver on the right represents the spine. At last, the pancreas is drawn as a black sphere (bottom right from the tumor).
\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{pictures/body}
\caption{Example body with tumor}
\label{model:body}
\end{figure}

Inside the body, the seeds are positioned randomly within the tumor voxels. When eradiating into the surrounding tissue, the seeds influence overlap and form a superposition. Hence, the dosis in a voxel is calculated as a sum of each seed by evaluating the dose-function up to a fixed limit, where the influence gets negligible (see chapter \ref{dose_function}).

@@

\section{Dose-Function}\label{dose_function}
The dose-distribution within the patient's body is the accumulation of all the radiation-seeds. Therefore, it is mandatory to know the radiation intensity of each seed. Equation \eqref{eq:dosefunction} shows the general approach to approximate the dose intensity function, whilst \eqref{eq:dosefunction2} is showing the used implementation of that equation.
\begin{equation}
\label{eq:dosefunction}
\dot{D}(r,\theta_{0})= S_{k}\Lambda \frac{G_{L}(r,\theta_{0})}{G_{L}(r_{0},\theta_{0})} g_{L}(r) F(r_{0},\theta_{0})
\end{equation}
\begin{equation}
\label{eq:dosefunction2}
\Rightarrow \ \dot{D}(r,\theta_{0})= S_{k}\Lambda \frac{G_{L}(r,\theta_{0})}{G_{L}(r_{0},\theta_{0})} g_{L}(r) \cdot 1
\end{equation}
\\
The following section will describe the used parameters:
\begin{description}
\item[Air-Kerma-Strength $S_{k}$]~\\
The $S_{k}$ describes the \textbf{k}inetic \textbf{e}nergy \textbf{r}eleased per unit \textbf{ma}ss called kerma. It is measured in $U$ ($1\ U =1 \ \frac{cGy \cdot m^{2}}{h} $)   
\item[Dose Rate Constant $\Lambda$]~\\
The dose rate constant is defined as the ratio of the dose rate at reference point $ P(r_{0},\theta_{0})  $ and the air-kerma-strength $S_{k}$. Therefore $ \Lambda \ = \ \frac{\dot{D}(r_{0},\theta_{0})}{S_{k}} $.

@@

 The population is the accumulation of all individuals (see \ref{pop}) $k$ denotes the number of individuals.
\begin{equation}
 \label{eq:population}
 Population \ \ p := \begin{pmatrix}
 i_{0} \\ \vdots \\ i_{k-1} 	
\end{pmatrix}   
 \end{equation}

@@

\paragraph{Treatment Range}
\label{para:treatment range}
As it is shown in chapter \ref{dose_function}, the distribution of the radiation decreases exponentially with increasing distance. This effect could be used to reduce the number of voxel being evaluated, as the influence of voxels relatively far away is negligibly small. Algorithm \ref{alg:scalefif} shows how that information is used. The region of interest (ROI) denotes the treatment range. \\

@@

\paragraph{Weights}
It seems hard to examine a specific sets of weights, that is applicable to any type of patient. The experiment has shown, that a higher weighted treatment has a much higher runtime and the convergency decreases.  Figure \ref{fig:weights} shows the difference between the goal dose and the actual dose for different weighted distributions.
\begin{figure}[hbtp]
\begin{tikzpicture}
  \begin{axis}[
  	name=one,
  	width=0.86 \columnwidth,
  	height=1.0in, 
  	scale only axis,
  	xmajorticks=false,
  	xtick={0,10,20,30,40},
  	ytick={-20,-10,...,10},
  	minor ytick={-20,-15,...,10},
  	legend style={anchor=north,at={(0.5,1.6)},font=\small},
  	legend columns=2,
    xmin=0,
    xmax=40,
  	ymax = 5,
  	ymin = -20,
    grid=both,
    y label style={at={(axis description cs:0,.5)},anchor=south},
    ylabel = {Diff. to GD [Gy]}
    ]
	\addplot [color=black,dashed,line width=1.0pt]
	table[row sep=crcr]{0 0\\
						10 -7.4061373\\
						20 -7.3660937\\
						30 -7.16107\\
						40 -7.5223036\\	};
	\addlegendentry{Bladder};

@@

\section{Linear Programming}
In linear programming an objective function and constraints are used to compute the optimal solution to a minimization or maximization problem, as described in [\cite{matprog}]. The problem consists of a certain number of variables and a feasible region, which is defined by the contraints. 
Equation \eqref{eq:LPMatrix} shows the standard form of a linear problem. The objective function, modeled with \textit{c},  is to be minimized, while the constraints, modeled with \textit{A} and \textit{b}, are satisfied. The used algorithm yields a vector $x$ for the optimal solution.
